Statistical Grand Slams

### 2023-12-15 Lahman Analysis/Formal Analysis/Draft1.qmd
Finished rough draft 2

Furthermore, it is logical to ask: why not consider both? Incorporating more predictors can often make a model more accurate at prediction; and both Fellingham and Fisher (2017) and this paper are concerned with maximizing predictive power of our respective models. The reason this paper does not consider both is because of multicollinearity. Multicollinearity is where two independent variables in a model are highly or perfectly correlated and the model is unable to determine which variable is responsible for the corresponding change in the depending variable. Without a doubt there is a strong correlation between PA's and AB's because one is a subset of the other. In fact, we can see that the correlation between the two is quite strong in @fig-BattingCor. If we included both PA's and AB's in our model, our model's alorithm Markov Chain Monte Carlo (MCMC) would have a difficult time determining which of the two variables is responsible for the changes in the dependent variable, HR's. For other generalized linaer models (GLM's) it becomes impossible to deal with mullticolinearity and either the model parameters need to be changed or the model is deemed imossible to calculate. Thus, we neglect PA's from our model both because theory stipulates it is an inferior predictor and used in tandem with our current predictor it would cause problems for the modeling software.

An interesting quirk of baseball is that, like soccer, stadiums differ in dimensions. This is not universal; baseball fields and stadiums must have the same measurements for their infields. Yet, for the most part baseball stadiums are allowed to configure their outfields as they please. This becomes important because most HR's are achieved by hitting the baseball over the outfield fence. Thus, fences closer to home plate generally lead to more HR's for all hitters, ceteris paribus. The non-standardized nature of stadiums' fences gives them a variety of unique shapes and ranging outfield sizes. For example, Fenway Park (1923-) is known to have the closest outfield wall with their "Green Monster" in left field being only 305 ft. from home plate. While the Polo Grounds (1891-1963) boasted one of the furthest walls from home plate at 483 ft. in center field. Stadiums also differ in location. Marlins Park (2012-2020) sits at 15 ft. above sea level while Coors Field (1995-) sits at 5183 ft. above sea level. These changes in eleveation allow the ball to fly farther in the air, and hence also impact HR rates. As illustrated there are not only vastly different dimensions that baseball fields can take on but also different environments which affect a player's ability to hit home runs. Thus, any model which attempts to model players' ability to hit HR's needs to consider ballpark factors.

@@

The parameter $\delta_p$ (which is $\xi_p$ in the actual paper) is the effect that playing in park $p$ has on the HR probability $\pi_{nip}$. As discussed in @sec-BaseBack, there is strong precedent to include such a term because the variance between ballpark dimension and enviornment can be drastic. Fellingham and Fisher (2017) estimate the effect that each park has on a player's ability to hit a HR and then add the associated effect to the players who played at that ballpark in a given year.

@@

Another questionable choice that Fellingham and Fisher (2017) make is that they estimated their model parameters using FORTRAN. Given the antiquity of FORTRAN, replication of Fellingham and Fisher (2017) is quite difficult. Not to mention that the 200,000 iterations makes the research so computationally expensive that it is nearly impossible for any other research team to replicate. This is disappointing because the scientific method relies on replication for the advancement of knowledge. Thus, our research seeks to make the algorithms used in this research more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language. Furthermore, our model will use less iterations and stronger priors so that other research teams can more easily repicate our findings.

@@

Given all the previous problems with Fellingham and Fisher (2017) mentioned above, there is an argument to be made that a simple generalized linear model (GLM) could be used instead of their complicated Bayesian semiparametric model. While it is true that a GLM could be used to predict HR's of a player it would not nearly be as accurate as a Bayesian model. This is because the greatest strength of using Bayesian modeling is that it allows for multilevel models to be created. That is, Bayesian models allow "learning" from clusters (sometimes known as "pools") of observations when there is not as much data available on a particular obersvation. Our model employs this method much more liberally than Fellingham and Fisher (2017). In turn, this should allow for us to make stronger inferences than Fellingham and Fisher (2017) because of this "learning".

@@

As discussed in @sec-BaseBack, there are many factors that influence player $n$'s underlying probability of hitting a $HR$ in a given $AB$. The first parameter which affects this probability, $\pi_{nip}$, is $\alpha_n$. In our model $\alpha_n$ can generally be thought of as the intercept term which is used to account for a player's innate ability to hit HR's. From our exploratory observations of the data, we noticed that the average (median or mean) proportion of AB's which resulted in HR's is about 0.03. This means that about 3% AB's result in HR's. This can be seen in @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall. On the logit scale, which is what our model relies on for estimation, a probability of 0.03 is about -3.5. We recognize that some players, for whatever reason, have better "innate" abilities to hit HR's than others. Therefore, we used a multilevel modeling technique for $\alpha_n$ to capture that player-to-player difference. To begin capturing this difference in $\alpha_n$ across players, we set the hyper-priors for $\alpha_n$ to be $\mu_0$ and $\sigma_0$. Where $\mu_0\sim Normal(-3.5,0.1)$ and $\sigma_0\sim Exponential(1)$. We set the mean of $\mu_0$ to be -3.5 because we estimate $\alpha_n$ on the logistic scale and we want it to be close to the average player's value. Since the parameters of $\alpha_n$ are distributions, we can think of $\alpha_n$ as a set of distributions, for the different values that either $\mu_0$ or $\sigma_0$ could take on. Further, the likelihood of the particular distribution that $\alpha_n$ takes on is depndent on the likelihood of either $\mu_0$ or $\sigma_0$ taking on a particular value. For example, $\mu_0=-3.0$ is extremely unlikely because it lies four standard deviations away from the mean of its normal distribution -3.5. Thus a distribution for $\alpha_n$ being $Normal(-3,\sigma_0)$ is highly unlikely for any value of $\sigma_0$. We chose $\mu_0$ to be normally distributed because there is precedent in Fellingham and Fisher (2017) to use normal distributions for their mean hyper-priors. Another reason it is easier to interpret normal distributions. We chose $\sigma_0$ to be exponentially distributed because the exponential distribution does not allow for negative values and the bulk of the possible values of $\sigma_0$ would be clustered near 0. Since we cannot have negative values for a standard deviation it seemed appropriate. This use of these hyper-priors is what makes the model appear to "learn". The model already comes in with certain expectations of what values $\alpha_n$ should take on, rather than simply coming into each new $n$ knowing nothing, which is what happens with more traditional GLM's. Finally, with this new set of distributions for $\alpha_n$ being the priors, our model constructs a posterior using Bayes theorem and the data on player $n$. It repeats this process for all 657 players that we considered for this research question to create a different $\alpha$ distribution for each player, $n$. Whatever value in $\alpha_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.

The next parameter we considered which affects $\pi_{nip}$ was $\beta_n$ which represents the effect centered age (the player's age minus 30) had on player $n$'s ability to hit HR's. This topic was briefly touched upon in @sec-BaseBack, where we discussed how a player's age can impact their their strategy and physical prowess. These ideas are well demonstrated in @fig-HR.By.Age.20to40. In @fig-HR.By.Age.20to40, one can easily see that, in general, players' ability to hit HR's improves over the course of their career and peaks when players reach about 30 years old. After this peak, a player's ability starts to decline for the rest of their career. We chose to center the the age of the all players at 30 for several reasons. The first was that roughly the mean and the median ages in the data was 30. Second, having a centered age makes interpreting $\beta_n$ easier. The value of $\beta_n$ represents how changes away from the centered age affects player $n$'s probability of hitting a HR, $\pi_{nip}$. In a normal GLM's, $\beta_n$ would not be able to adapt to each player, $n$, as it does in our model. It would only be able to consider how aging affects the average player's HR probability and then apply that effect to all players. This implicitly assumes that each player is affected identically by changes in age. In contrast, we employ multilevel modeling techniques to our model where $\beta_n\sim Normal(\mu_1,\sigma_1)$, $\mu_1\sim Normal(0,0.1)$, and $\sigma_1\sim Exponential(10)$. Similar to understanding $\alpha_n$, we can interpret $\beta_n$ as a collection of distributions corresponding to the different values of its hyper-priors $\mu_1$ and $\sigma_1$ and their corresponding likelihoods. We chose a normal distribution for $\mu_1$ for reasons similar to $\mu_0$. It makes interpretation easier and there is precedent in the literature. Furthermore, we chose it to have a mean of 0 because we assume that there is only a slighly poitive increase in HR probability year over year, which is close enough to zero that we let the data speak for itself. Again, for similar reasons to $\sigma_0$ we chose $\sigma_1$ to be exponentially distriubted. This ensured that $\sigma_1$ would not be negative and that the bulk of the possible values of $\sigma_1$ would be clustered near 0. With the distributions calculated for $\beta_n$ we can now use $\beta_n$ as the prior and combine it with the player data to generate the posterior using Bayes theorem. We repeat this process for all 657 players considered to get a unique distribution for each player $\beta_n$. This means that each player, $n$ has a unique aging coefficient which represents how aging effects their HR probability. Whatever value in $\beta_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.

@@

Similar to Fellingham and Fisher (2017), we have used the Lahman data set for player statistics. While Lahman has data on a wide range of MLB realted topics, we decided to focus on a singular data sheet. We focused on the "Batting" data set in Lahman which describes player-year-team batting statistics. That is, each entry in Batting is for a player $n$ who played in year $i$ for team $t$. This allows multiple entries for the same year for the same player if they were traded, which is important for our analysis because different teams implies different home ballparks.

In order to be considered in our analysis a player had to be in Batting and have played for at least 6 seasons from 1973-2019. We chose these cut-offs because in 1973 the American League allowed the Designated Hitter, signaling that there would be an increased specialization in labor for hitting. We chose 2019 as the other cut-off because 2020 was the COVID season which was drastically shortened and very different compared to almost all other previous seasons. Furthermore, in order to be considered in our analysis players had to have had at least 50 AB's across those 6 seasons played. The effect of these cut-offs was that we were left with players who had longer careers and who hit HR's. There were many players dropped who never hit HR's because of their short stints which made prediction hard and prediction for these playes did not seem fruitful given their general inconsequential careers.

@@

Consider player $n=651$. That is, consider the player Robin Yount. In @fig-p651Act we can see how Yount's actual HR probability has changed over the years that he played. The x-axis of @fig-p651Act describes Yount's deviation from the centered age, where the centered age value of 0 represents a player being 30 years old. And the y-axis represents the actual HR probability of Yount. Thus, the point (0, 3.25) tells us that Yount's actual HR probability was 3.25% when he was 30 years old. Similarly, the points (-10, 1.5) and (6, 2) represent Yount's actual HR probability being 1.5% when he was 20 years old and 2% when he was 36 years old, respectively. Included in @fig-p651Act is a blue trend line that snakes through the data. This line does not represent anything meaningful, it was included simply to make seeing and interpreting the trend present in the scatter-plot easier. To that end, we can see in @fig-p651Act that during the first 10 years of his career, Yount saw a steady increase in his actual HR probability until he was 28 years old. From when he was 29 years old until he was about 35 years old, there some fluctuation in his HR probability by overall it stayed fairly consistent at around 2.5-3%. Finally, in the last three years of his career we can see Yount's HR probability decline to about 1.75% when he was 38 years old. Overall, Yount appears to have a typical career compared to what the told our model to expect. There was a steady increase in probability until he was about 30, followed by a decline afterwards as he got older. The shape of the trend-line through the scatter-plot follows a vaguely parabolic shape which is what we would expect to see. Thus, Yount will be a good benchmark for the model to see how it predicts "typical" players.

In @fig-p651Pred we can examine how our model predicted Yount's HR probability, $\pi$, to change over his career. The x-axis is the same as it is in @fig-p651Act, where it describes Yount's deviation from the centered age (which is 30). The y-axis describes Yount's predicted HR probability $\pi$. An important note is that the y-axis is not measured on the logistic scale. We calculated it on the logistic scale but then converted it back to the normal scale, and use the normal scale in the figure, to prevent any confusion. Interpreting the points in @fig-p651Pred is very similar to the interpretation of points for @fig-p651Act. Each point in @fig-p651Pred represents the model's prediction of Yount's $\pi$ for a centered age value. For example, the point (0, 3.25) represents the model predicting that Yount's $\pi=3.25$ when he is 30 years old, which was very close to the actual value. Similarly, the points (-10, 1) and (6, 2) represent our model's prediction that Yount's $\pi=1$ when he is 20 years old and $\pi=2$ when he is 36 years old, respectively, which is also very close to what we saw in @fig-p651Act. Included in @fig-p651Pred is a red trend line for the same reasons mentioned in discussing @fig-p651Act. The red trend line represents the overall trend of the data in the scatter-plot. It is not nearly as necessary here, because the data is so clearly clustered in a parabolic pattern, but it is helpful in future plots. We can see in @fig-p651Pred that the predictions from our model follow closely to what we told it to expect. There is an increase in HR probability, $\pi$, for Yount from when he is 18 to when he is 32. This is quite close to what we see in the figure representing his actual HR probability. Then we see a sharp decline in predicted HR probability for Yount as he ages past 32 until he retires at the age of 38.

@@

Consider player $n=582$. That is, consider the player Pat Tabler. In @fig-p582Act we can see how Tabler's actual HR probability has changed over the years that he played. The x-axis of @fig-p582Act describes Tabler's deviation from the centered age, where the centered age value of 0 represents a player being 30 years old. And the y-axis represents the actual HR probability of Tabler. Thus, the point (-1, 2) tells us that Tabler's actual HR probability was 2% when he was 29 years old. In fact, the interpretation of @fig-p582Act is exactly the same as it was for @fig-p651Act, expect the probabilities are for Pat Tabler instead of Robin Yount. Observant readers may have noticed that there are two data points when $x=0$ and when $x=2$. This is because Pat Tabler was traded twice, once when he was 30 and another time when he was 32 years old. Since he was traded, it meant that he had two different home ballparks for that season. Thus, he had two different HR probabilities for that year with each representing his HR probability for a specific home ballpark. This "double observation" does not impact our impact our model at all. It is simply a quirk of the data. Included in @fig-p582Act is a blue trend line that snakes through the data. This line does not represent anything meaningful, it was included simply to make seeing and interpreting the trend present in the scatter-plot easier. To that end, we can see in @fig-p582Act that for the first 7 years of his career Tabler saw an increase in his actual HR probability, which lasted until he was 29 years old. Tabler's average HR probability at this time was about 1.5% during this period. Every year after he turned 29, with the exception of when he was 32, he performed starkly worse than he did in the beginning of his career. The average of these points being around 0.5%. The only outlier in this period was the season where he was 32, where he was traded to the New York Mets and had a higher HR probability for them that year then he did with the Kansas City Royals. However, this higher HR probability was probably due to sample size. He had the same amount of HR's with both teams but more than 4 times as many AB's with the Royals compared with the Mets. Overall, we can see that Tabler's career was different than Yount's in a couple key ways. The first is that Tabler stays well below the average HR probability for all of his career, whereas Yount says very close to it. The second way is that Tabler peaks before he is 30 and does so bad for the final years of his career that there is not much decline. Thus, it will be interesting to see how our adapts to this very interesting and unique player.

In @fig-p582Pred we can examine how our model predicted Tabler's HR probability, $\pi$, to change over his career. The x-axis is the same as it is in @fig-p582Act, where it describes Tabler's deviation from the centered age (which is 30). The y-axis describes Tabler's predicted HR probability $\pi$. An important note is that the y-axis is not measured on the logistic scale. We calculated it on the logistic scale but then converted it back to the normal scale, and use the normal scale in the figure, to prevent any confusion. Interpreting the points in @fig-p582Pred is very similar to the interpretation of points for @fig-p582Act. Each point in @fig-p582Pred represents the model's prediction of Tabler's $\pi$ for a centered age value. For example, the point (-1, 1.55) represents the model predicting that Tabler's $\pi=1.55$ when he is 29 years old. Similarly, the points (-4, 1.2) and (4, 0.8) represent our model's prediction that Tabler's $\pi=1.2$ when he is 26 years old and $\pi=0.8$ when he is 34 years old, respectively. However, the $\pi$ values were not nearly as accurate for Tabler as they were for Yount. We can see that the high variation of the actual HR probabilities has forced our model rely more on the priors than on the actual data. That is, we can see our estimates for $\pi$ are often much closer to the average HR rate across players, which is 3%, than they are to Tabler's specific data because Tabler is such an outlier. Included in @fig-p582Pred is a red trend line for the same reasons mentioned in discussing @fig-p582Act. The red trend line represents the overall trend of the data in the scatter-plot. Coming into this, the worry was that our model would rely too heavily on its priors and not be able to update its parabolic shape to match the strange career that Tabler had. However, after quickly examining the red line anyone can see that it has clearly updated to fit Tabler's data much better than @fig-p651Pred. That is, we can see that the red line from @fig-p582Pred matches the blue line in @fig-p582Act fairly well, demonstrating that the model updates itself based on the data it is fed. We can see that red line captures Tabler's beginning career well, with a slight increase in HR probability until he is 29. This is then followed by a steep and drastic drop in predicted HR rate for all years including and past when Tabler is 30 years old. Thus, although the actual proportions themselves do not line up as well as we may have hoped, we can see that the model does match trends well, which is a good sign. Thus, it will be interesting to see how different extreme player trajectories challenge the model.

@@

See @fig-p420Act and @fig-p420Pred.

>>>>

Furthermore, it is logical to ask: why not consider both? Incorporating more predictors can often make a model more accurate at prediction; and both Fellingham and Fisher (2017) and this paper are concerned with maximizing predictive power of our respective models. The reason this paper does not consider both is because of multicollinearity. Multicollinearity is where two independent variables in a model are highly or perfectly correlated and the model is unable to determine which variable is responsible for the corresponding change in the depending variable. Without a doubt there is a strong correlation between PA's and AB's because one is a subset of the other. In fact, we can see that the correlation between the two is quite strong in @fig-BattingCor. If we included both PA's and AB's in our model, our model's algorithm Markov Chain Monte Carlo (MCMC) would have a difficult time determining which of the two variables is responsible for the changes in the dependent variable, HR's. For other generalized linear models (GLM's) it becomes impossible to deal with multicollinearity and either the model parameters need to be changed or the model is deemed impossible to calculate. Thus, we neglect PA's from our model both because theory stipulates it is an inferior predictor and used in tandem with our current predictor it would cause problems for the modeling software.

An interesting quirk of baseball is that, like soccer, stadiums differ in dimensions. This is not universal; baseball fields and stadiums must have the same measurements for their infields. Yet, for the most part baseball stadiums are allowed to configure their outfields as they please. This becomes important because most HR's are achieved by hitting the baseball over the outfield fence. Thus, fences closer to home plate generally lead to more HR's for all hitters, ceteris paribus. The non-standardized nature of stadiums' fences gives them a variety of unique shapes and ranging outfield sizes. For example, Fenway Park (1923-) is known to have the closest outfield wall with their "Green Monster" in left field being only 305 ft. from home plate. While the Polo Grounds (1891-1963) boasted one of the furthest walls from home plate at 483 ft. in center field. Stadiums also differ in location. Marlins Park (2012-2020) sits at 15 ft. above sea level while Coors Field (1995-) sits at 5183 ft. above sea level. These changes in elevation allow the ball to fly farther in the air, and hence also impact HR rates. As illustrated there are not only vastly different dimensions that baseball fields can take on but also different environments which affect a player's ability to hit home runs. Thus, any model which attempts to model players' ability to hit HR's needs to consider ballpark factors.

@@

The parameter $\delta_p$ (which is $\xi_p$ in the actual paper) is the effect that playing in park $p$ has on the HR probability $\pi_{nip}$. As discussed in @sec-BaseBack, there is strong precedent to include such a term because the variance between ballpark dimension and environment can be drastic. Fellingham and Fisher (2017) estimate the effect that each park has on a player's ability to hit a HR and then add the associated effect to the players who played at that ballpark in a given year.

@@

Another questionable choice that Fellingham and Fisher (2017) make is that they estimated their model parameters using FORTRAN. Given the antiquity of FORTRAN, replication of Fellingham and Fisher (2017) is quite difficult. Not to mention that the 200,000 iterations makes the research so computationally expensive that it is nearly impossible for any other research team to replicate. This is disappointing because the scientific method relies on replication for the advancement of knowledge. Thus, our research seeks to make the algorithms used in this research more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language. Furthermore, our model will use less iterations and stronger priors so that other research teams can more easily replicate our findings.

@@

Given all the previous problems with Fellingham and Fisher (2017) mentioned above, there is an argument to be made that a simple generalized linear model (GLM) could be used instead of their complicated Bayesian semiparametric model. While it is true that a GLM could be used to predict HR's of a player it would not nearly be as accurate as a Bayesian model. This is because the greatest strength of using Bayesian modeling is that it allows for multilevel models to be created. That is, Bayesian models allow "learning" from clusters (sometimes known as "pools") of observations when there is not as much data available on a particular observation. Our model employs this method much more liberally than Fellingham and Fisher (2017). In turn, this should allow for us to make stronger inferences than Fellingham and Fisher (2017) because of this "learning".

@@

As discussed in @sec-BaseBack, there are many factors that influence player $n$'s underlying probability of hitting a $HR$ in a given $AB$. The first parameter which affects this probability, $\pi_{nip}$, is $\alpha_n$. In our model $\alpha_n$ can generally be thought of as the intercept term which is used to account for a player's innate ability to hit HR's. From our exploratory observations of the data, we noticed that the average (median or mean) proportion of AB's which resulted in HR's is about 0.03. This means that about 3% AB's result in HR's. This can be seen in @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall. On the logit scale, which is what our model relies on for estimation, a probability of 0.03 is about -3.5. We recognize that some players, for whatever reason, have better "innate" abilities to hit HR's than others. Therefore, we used a multilevel modeling technique for $\alpha_n$ to capture that player-to-player difference. To begin capturing this difference in $\alpha_n$ across players, we set the hyper-priors for $\alpha_n$ to be $\mu_0$ and $\sigma_0$. Where $\mu_0\sim Normal(-3.5,0.1)$ and $\sigma_0\sim Exponential(1)$. We set the mean of $\mu_0$ to be -3.5 because we estimate $\alpha_n$ on the logistic scale and we want it to be close to the average player's value. Since the parameters of $\alpha_n$ are distributions, we can think of $\alpha_n$ as a set of distributions, for the different values that either $\mu_0$ or $\sigma_0$ could take on. Further, the likelihood of the particular distribution that $\alpha_n$ takes on is dependent on the likelihood of either $\mu_0$ or $\sigma_0$ taking on a particular value. For example, $\mu_0=-3.0$ is extremely unlikely because it lies four standard deviations away from the mean of its normal distribution -3.5. Thus a distribution for $\alpha_n$ being $Normal(-3,\sigma_0)$ is highly unlikely for any value of $\sigma_0$. We chose $\mu_0$ to be normally distributed because there is precedent in Fellingham and Fisher (2017) to use normal distributions for their mean hyper-priors. Another reason it is easier to interpret normal distributions. We chose $\sigma_0$ to be exponentially distributed because the exponential distribution does not allow for negative values and the bulk of the possible values of $\sigma_0$ would be clustered near 0. Since we cannot have negative values for a standard deviation it seemed appropriate. This use of these hyper-priors is what makes the model appear to "learn". The model already comes in with certain expectations of what values $\alpha_n$ should take on, rather than simply coming into each new $n$ knowing nothing, which is what happens with more traditional GLM's. Finally, with this new set of distributions for $\alpha_n$ being the priors, our model constructs a posterior using Bayes theorem and the data on player $n$. It repeats this process for all 657 players that we considered for this research question to create a different $\alpha$ distribution for each player, $n$. Whatever value in $\alpha_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.

The next parameter we considered which affects $\pi_{nip}$ was $\beta_n$ which represents the effect centered age (the player's age minus 30) had on player $n$'s ability to hit HR's. This topic was briefly touched upon in @sec-BaseBack, where we discussed how a player's age can impact their their strategy and physical prowess. These ideas are well demonstrated in @fig-HR.By.Age.20to40. In @fig-HR.By.Age.20to40, one can easily see that, in general, players' ability to hit HR's improves over the course of their career and peaks when players reach about 30 years old. After this peak, a player's ability starts to decline for the rest of their career. We chose to center the the age of the all players at 30 for several reasons. The first was that roughly the mean and the median ages in the data was 30. Second, having a centered age makes interpreting $\beta_n$ easier. The value of $\beta_n$ represents how changes away from the centered age affects player $n$'s probability of hitting a HR, $\pi_{nip}$. In a normal GLM's, $\beta_n$ would not be able to adapt to each player, $n$, as it does in our model. It would only be able to consider how aging affects the average player's HR probability and then apply that effect to all players. This implicitly assumes that each player is affected identically by changes in age. In contrast, we employ multilevel modeling techniques to our model where $\beta_n\sim Normal(\mu_1,\sigma_1)$, $\mu_1\sim Normal(0,0.1)$, and $\sigma_1\sim Exponential(10)$. Similar to understanding $\alpha_n$, we can interpret $\beta_n$ as a collection of distributions corresponding to the different values of its hyper-priors $\mu_1$ and $\sigma_1$ and their corresponding likelihoods. We chose a normal distribution for $\mu_1$ for reasons similar to $\mu_0$. It makes interpretation easier and there is precedent in the literature. Furthermore, we chose it to have a mean of 0 because we assume that there is only a slightly positive increase in HR probability year over year, which is close enough to zero that we let the data speak for itself. Again, for similar reasons to $\sigma_0$ we chose $\sigma_1$ to be exponentially distributed. This ensured that $\sigma_1$ would not be negative and that the bulk of the possible values of $\sigma_1$ would be clustered near 0. With the distributions calculated for $\beta_n$ we can now use $\beta_n$ as the prior and combine it with the player data to generate the posterior using Bayes theorem. We repeat this process for all 657 players considered to get a unique distribution for each player $\beta_n$. This means that each player, $n$ has a unique aging coefficient which represents how aging effects their HR probability. Whatever value in $\beta_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.

@@

Similar to Fellingham and Fisher (2017), we have used the Lahman data set for player statistics. While Lahman has data on a wide range of MLB related topics, we decided to focus on a singular data sheet. We focused on the "Batting" data set in Lahman which describes player-year-team batting statistics. That is, each entry in Batting is for a player $n$ who played in year $i$ for team $t$. This allows multiple entries for the same year for the same player if they were traded, which is important for our analysis because different teams implies different home ballparks.

In order to be considered in our analysis a player had to be in Batting and have played for at least 6 seasons from 1973-2019. We chose these cut-offs because in 1973 the American League allowed the Designated Hitter, signaling that there would be an increased specialization in labor for hitting. We chose 2019 as the other cut-off because 2020 was the COVID season which was drastically shortened and very different compared to almost all other previous seasons. Furthermore, in order to be considered in our analysis players had to have had at least 50 AB's across those 6 seasons played. The effect of these cut-offs was that we were left with players who had longer careers and who hit HR's. There were many players dropped who never hit HR's because of their short stints which made prediction hard and prediction for these players did not seem fruitful given their general inconsequential careers.

@@

Consider player $n=651$. That is, consider the player Robin Yount. In @fig-p651Act we can see how Yount's actual HR probability has changed over the years that he played. The x-axis of @fig-p651Act describes Yount's deviation from the centered age, where the centered age value of 0 represents a player being 30 years old. And the y-axis represents the actual HR probability of Yount. Thus, the point (0, 3.25) tells us that Yount's actual HR probability was 3.25% when he was 30 years old. Similarly, the points (-10, 1.5) and (6, 2) represent Yount's actual HR probability being 1.5% when he was 20 years old and 2% when he was 36 years old, respectively. Included in @fig-p651Act is a blue trend line that snakes through the data. This line does not represent anything meaningful, it was included simply to make seeing and interpreting the trend present in the scatter-plot easier. To that end, we can see in @fig-p651Act that during the first 10 years of his career, Yount saw a steady increase in his actual HR probability until he was 28 years old. From when he was 29 years old until he was about 35 years old, there some fluctuation in his HR probability by overall it stayed fairly consistent at around 2.5-3%. Finally, in the last three years of his career we can see Yount's HR probability decline to about 1.75% when he was 38 years old. Overall, Yount appears to have a typical career compared to what we told our model to expect. There was a steady increase in probability until he was about 30, followed by a decline afterwards as he got older. The shape of the trend-line through the scatter-plot follows a vaguely parabolic shape which is what we would expect to see. Thus, Yount will be a good benchmark for the model to see how it predicts "typical" players.

In @fig-p651Pred we can examine how our model predicted Yount's HR probability, $\pi$, to change over his career. The x-axis is the same as it is in @fig-p651Act, where it describes Yount's deviation from the centered age (which is 30). The y-axis describes Yount's predicted HR probability $\pi$. An important note is that the y-axis is not measured on the logistic scale. We calculated it on the logistic scale but then converted it back to the normal scale, and use the normal scale in the figure, to prevent any confusion. Interpreting the points in @fig-p651Pred is very similar to the interpretation of points for @fig-p651Act. Each point in @fig-p651Pred represents the model's prediction of Yount's $\pi$ for a centered age value. For example, the point (0, 3.25) represents the model predicting that Yount's $\pi=3.25$ when he is 30 years old, which was very close to the actual value. Similarly, the points (-10, 1) and (6, 2) represent our model's prediction that Yount's $\pi=1$ when he is 20 years old and $\pi=2$ when he is 36 years old, respectively, which is also very close to what we saw in @fig-p651Act. Included in @fig-p651Pred is a red trend line for the same reasons mentioned in discussing @fig-p651Act. The red trend line represents the overall trend of the data in the scatter-plot. It is not nearly as necessary here, because the data is so clearly clustered in a parabolic pattern, but it is helpful in future plots. We can see in @fig-p651Pred that the predictions from our model follow closely to what we told it to expect. There is an increase in HR probability, $\pi$, for Yount from when he was 18 to when he was 32. This is quite close to what we see in the figure representing his actual HR probability. Then we see a sharp decline in predicted HR probability for Yount as he ages past 32 until he retires at the age of 38.

@@

Consider player $n=582$. That is, consider the player Pat Tabler. In @fig-p582Act we can see how Tabler's actual HR probability has changed over the years that he played. The x-axis of @fig-p582Act describes Tabler's deviation from the centered age, where the centered age value of 0 represents a player being 30 years old. And the y-axis represents the actual HR probability of Tabler. Thus, the point (-1, 2) tells us that Tabler's actual HR probability was 2% when he was 29 years old. In fact, the interpretation of @fig-p582Act is exactly the same as it was for @fig-p651Act, expect the probabilities are for Pat Tabler instead of Robin Yount. Observant readers may have noticed that there are two data points when $x=0$ and when $x=2$. This is because Pat Tabler was traded twice, once when he was 30 and another time when he was 32 years old. Since he was traded, it meant that he had two different home ballparks for that season. Thus, he had two different HR probabilities for that year with each representing his HR probability for a specific home ballpark. This "double observation" does not impact our impact our model at all. It is simply a quirk of the data.

Included in @fig-p582Act is a blue trend line that snakes through the data. This line does not represent anything meaningful, it was included simply to make seeing and interpreting the trend present in the scatter-plot easier. To that end, we can see in @fig-p582Act that for the first 7 years of his career Tabler saw an increase in his actual HR probability, which lasted until he was 29 years old. Tabler's average HR probability at this time was about 1.5% during this period. Every year after he turned 29, with the exception of when he was 32, he performed starkly worse than he did in the beginning of his career. The average of these points being around 0.5%. The only outlier in this period was the season where he was 32, where he was traded to the New York Mets and had a higher HR probability for them that year then he did with the Kansas City Royals. However, this higher HR probability was probably due to sample size. He had the same amount of HR's with both teams but more than 4 times as many AB's with the Royals compared with the Mets. Overall, we can see that Tabler's career was different than Yount's in a couple key ways. The first is that Tabler stays well below the average HR probability for all of his career, whereas Yount says very close to it. The second way is that Tabler peaks before he is 30 and does so bad for the final years of his career that there is not much decline. Thus, it will be interesting to see how our adapts to this very interesting and unique career.

In @fig-p582Pred we can examine how our model predicted Tabler's HR probability, $\pi$, to change over his career. The x-axis is the same as it is in @fig-p582Act, where it describes Tabler's deviation from the centered age (which is 30). The y-axis describes Tabler's predicted HR probability $\pi$. An important note is that the y-axis is not measured on the logistic scale. We calculated it on the logistic scale but then converted it back to the normal scale, and use the normal scale in the figure, to prevent any confusion. Interpreting the points in @fig-p582Pred is very similar to the interpretation of points for @fig-p582Act. Each point in @fig-p582Pred represents the model's prediction of Tabler's $\pi$ for a centered age value. For example, the point (-1, 1.55) represents the model predicting that Tabler's $\pi=1.55$ when he is 29 years old. Similarly, the points (-4, 1.2) and (4, 0.8) represent our model's prediction that Tabler's $\pi=1.2$ when he is 26 years old and $\pi=0.8$ when he is 34 years old, respectively. However, the $\pi$ values were not nearly as accurate for Tabler as they were for Yount. We can see that the high variation of the actual HR probabilities has forced our model rely more on the priors than on the actual data. That is, we can see our estimates for $\pi$ are often much closer to the average HR rate across players, which is 3%, than they are to Tabler's specific data because Tabler is such an outlier.

Included in @fig-p582Pred is a red trend line for the same reasons mentioned in discussing @fig-p582Act. The red trend line represents the overall trend of the data in the scatter-plot. Coming into this, the worry was that our model would rely too heavily on its priors and not be able to update its parabolic shape to match the strange career that Tabler had. However, after quickly examining the red line anyone can see that it has clearly updated to fit Tabler's data. That is, we can see that the red line from @fig-p582Pred matches the blue line in @fig-p582Act fairly well, demonstrating that the model updates itself based on the data it is fed. We can see that red line captures Tabler's beginning career well, with a slight increase in HR probability until he is 29. This is then followed by a steep and drastic drop in predicted HR rate for all years including and past when Tabler is 30 years old. Thus, although the actual proportions themselves do not line up as well as we may have hoped, we can see that the model does match trends well, which is a good sign. Thus, it will be interesting to see how different extreme player trajectories challenge the model.

@@

Consider player $n=420$. That is, consider the player Kendrys Morales. In @fig-p420Act we can see how Morales' actual HR probability has changed over the years that he played. The x-axis of @fig-p420Act describes Morales' deviation from the centered age, where the centered age value of 0 represents a player being 30 years old. And the y-axis represents the actual HR probability of Morales. Thus, the point (0, 4.25) tells us that Morales' actual HR probability was 4.25% when he was 30 years old. In fact, the interpretation of @fig-p420Act is exactly the same as it was for both @fig-p651Act and @fig-p582Act, expect the probabilities are for Kendrys Morales instead of Pat Tabler or Robin Yount. Observant readers may have noticed that there are two data points when $x=1$ and when $x=6$. This is because Morales was traded twice, once when he was 31 and another time when he was 36 years old. Since he was traded, it meant that he had two different home ballparks for that season. Thus, he had two different HR probabilities for that year with each representing his HR probability for a specific home ballpark. This "double observation" does not impact our impact our model at all. It is simply a quirk of the data.
